A Comparative Study of Game-Theoretic and Reinforcement Learning Approaches for Jam-Resilient Wireless Networks in IoT

### 2025-05-29 paper/main.tex
About done

\graphicspath{{figures/}} % Tells LaTeX to look in the 'figures' subdirectory

@@

%======================== Title and Authors ========================

\title{A Comparative Study of Game-Theoretic and Reinforcement Learning Approaches for Jam-Resilient Wireless Networks in IoT}

@@

While simple techniques like channel hopping offer limited protection against static jammers, they fall short against adaptive adversaries. To address this, game-theoretic models have been proposed to anticipate attacker behavior through strategic reasoning \cite{jia2022game}. However, such models rely on predefined payoff structures and may underperform when network dynamics evolve unpredictably. On the other hand, reinforcement learning (RL) offers adaptive defense via trial-and-error policy optimization \cite{tang2024adaptive}, but it can require substantial exploration and may converge to suboptimal strategies—especially in noisy or sparse-reward environments.

@@

%======================== 2. Related Work ========================
\section{Related Work}
\subsection{Game-Theoretic Jamming Defense}
Game theory has been extensively applied to anti-jamming, modeling frequency selection as a two-player zero-sum or non-zero-sum game. Early work by Axell et al. \cite{axell2012detection} proposes Nash equilibrium strategies for frequency hopping, while Chen and Tong \cite{chen2019anti} incorporate energy constraints. Bayesian games extend this framework to handle incomplete information about attacker capabilities, improving robustness \cite{alpcan2007game,khouzani2012evolving}. Recent advances by Jia et al. \cite{jia2022game} demonstrate game-theoretic learning approaches for anti-jamming in wireless networks, while Singh and Gupta \cite{resource2023slicing} explore game theory applications in B5G resource allocation. However, the computational complexity grows rapidly with network size and action space.

\subsection{Reinforcement Learning for Security}
Reinforcement learning offers a dynamic alternative, enabling agents to learn channel-selection policies without explicit attacker modeling. Peng et al. \cite{peng2017anti} apply Q-Learning to anti-jamming, achieving significant throughput improvements, while Luo et al. \cite{luo2018anti} leverage deep RL for spectrum sharing. Rahman et al. \cite{rahman2023rlbased} propose a multi-agent game-theoretic framework for RL-based attack detection in IoT networks, demonstrating the potential of hybrid approaches. Recent work by Tang et al. \cite{tang2024adaptive} shows adaptive anti-jamming strategies for UAV-aided IoT communications using reinforcement learning. Hybrid approaches combining game theory and RL have emerged, demonstrating enhanced performance \cite{zhao2020hybrid,li2021game}.

\subsection{Comparative Analyses}
Comparisons between game-theoretic and RL-based defenses are scarce. Zhao et al. \cite{zhao2020hybrid} evaluate hybrid schemes but focus on throughput only. Comprehensive surveys \cite{survey2022jamming} highlight the need for systematic comparisons across multiple performance dimensions. Our study distinguishes itself by evaluating multiple metrics—including payoffs, detection rates, and health—under a unified experimental setup.

%======================== 3. Methodology ========================
\section{Methodology}
\subsection{Simulation Environment}
We simulate interactions on four distinct network topologies to reflect diverse IoT deployment patterns: (i) Erd\H{o}s--R\'enyi random graphs, (ii) Star topologies, (iii) Ring networks, and (iv) Small-World graphs generated via the Watts–Strogatz model. Each topology is instantiated for three different network sizes: $n = 10$, $n = 20$, and $n = 50$ nodes. For Erd\H{o}s--R\'enyi graphs, the connectivity probability is set to $p=0.4$ to ensure moderate density.

@@

Q-values are updated using the standard temporal-difference learning rule:
\[
Q(s_t, a_t) \leftarrow Q(s_t, a_t) + \alpha \left[ r_t + \gamma \max_{a'} Q(s_{t+1}, a') - Q(s_t, a_t) \right],
\]
with learning rate $\alpha = 0.1$, discount factor $\gamma = 0.9$, and $\epsilon$-greedy exploration strategy initialized with $\epsilon_0 = 0.5$ (decaying to $\epsilon_{\min} = 0.01$ at a rate of 0.998).

@@

All simulations were conducted on a personal laptop equipped with an AMD Ryzen 5 5600H CPU and 16 GB RAM. No parallel or GPU acceleration was used. While the defense strategies are intended for deployment in resource-constrained IoT networks, the experiments were run in a simulated environment to evaluate algorithmic performance and comparative trade-offs under controlled conditions.

@@

\subsection{Convergence Analysis}
Figure~\ref{fig:conv_db} illustrates defender payoff trajectories over time for Q-Learning and Bayesian strategies. Q-Learning defenders show rapid early improvement due to exploratory behavior, but plateau after step 150, often converging to locally optimal policies. In contrast, Bayesian defenders exhibit a slower but more stable convergence pattern, achieving higher long-term payoff with minimal inter-trial variance. This behavior reflects the advantage of equilibrium-driven decision-making under uncertainty.

@@

\subsection{Boxplot Comparisons}
Figure~\ref{fig:def_box} shows the distribution of final defender payoffs at $T=300$ across defender strategies (hue indicates attacker type). Bayesian defenders exhibit tighter payoff distributions (IQR = 0.2) compared to Q-Learning (IQR = 0.35), indicating more stable and predictable behavior. Additionally, the median payoff for Bayesian strategies is higher in all attacker pairings, underscoring their robustness against adversarial variability.

@@

Our comparative results indicate that Bayesian game-theoretic defenses offer consistently robust performance with low variance, making them well-suited for IoT applications that require predictable security guarantees. Among all evaluated metrics, the most pronounced improvement is observed in detection rate, where Bayesian defenders outperform Q-Learning counterparts by approximately 7\%. This reliability stems from the model's use of prior beliefs and equilibrium reasoning, which ensures stability even under uncertainty.

However, the computational overhead of belief updates and equilibrium computation may hinder real-time deployment in larger or more dynamic networks, particularly on constrained edge devices. In contrast, Q-Learning defenders demonstrate fast initial adaptation to attacker behavior, achieving higher early-stage payoffs. Yet, their long-term performance often degrades due to convergence to local optima, resulting in up to 10\% lower final payoffs compared to Bayesian strategies.

@@

Although our simulations were conducted on a general-purpose laptop, the algorithms were intentionally designed with computational efficiency in mind. For real-world IoT deployments, further evaluation on hardware platforms such as Raspberry Pi or ESP32 would help assess energy usage, memory footprint, and real-time feasibility. Measuring latency per decision cycle and reward computation would provide critical insights into deployment viability.

@@

The findings highlight a core trade-off between stability and adaptability, suggesting that future systems may benefit from hybrid designs that combine the strengths of both approaches. The modular simulator and all experimental artifacts have been made publicly available to facilitate reproducibility and further exploration.

>>>>

\graphicspath{{figures/}}

@@

%======================== Title and Authors ========================
\title{A Comparative Study of Game-Theoretic and Reinforcement Learning Approaches for Jam-Resilient Wireless Networks in IoT}

@@

Recent studies have emphasized the rising complexity of jamming techniques in evolving communication ecosystems like 5G/6G and industrial IoT, necessitating more intelligent countermeasures \cite{lohan2024ai, bensaad2023jamming}. While simple techniques like channel hopping offer limited protection against static jammers, they fall short against adaptive adversaries. To address this, game-theoretic models have been proposed to anticipate attacker behavior through strategic reasoning \cite{jia2022game}. However, such models rely on predefined payoff structures and may underperform when network dynamics evolve unpredictably. On the other hand, reinforcement learning (RL) offers adaptive defense via trial-and-error policy optimization \cite{tang2024adaptive}, but it can require substantial exploration and may converge to suboptimal strategies—especially in noisy or sparse-reward environments.

@@

%======================== 2. Related Work ========================
\section{Related Work}

\subsection{Game-Theoretic Jamming Defense}
Game theory has been extensively applied to anti-jamming, modeling frequency selection as a two-player zero-sum or non-zero-sum game. Early work by Axell et al. \cite{axell2012detection} proposes Nash equilibrium strategies for frequency hopping, while Chen and Tong \cite{chen2019anti} incorporate energy constraints. Bayesian games extend this framework to handle incomplete information about attacker capabilities, improving robustness \cite{alpcan2007game,khouzani2012evolving}. Recent developments include Stackelberg models and healthcare-oriented applications, such as Gouissem et al.'s Nash game-based power allocation strategy for jam-resilient healthcare IoT \cite{gouissem2022nash} and Imran et al.'s Stackelberg game leveraging DSSS for anti-jamming in cognitive radio networks \cite{imran2024stackelberg}. Barkatsa et al. \cite{barkatsa2024bayesian} propose a Bayesian game-theoretic uplink allocation method tailored for federated IoT under jamming scenarios. Jia et al. \cite{jia2022game} further demonstrate game-theoretic learning approaches for anti-jamming in wireless networks, while Singh and Gupta \cite{resource2023slicing} explore game theory applications in B5G resource allocation. However, the computational complexity grows rapidly with network size and action space.

\subsection{Reinforcement Learning for Security}
Reinforcement learning offers a dynamic alternative, enabling agents to learn channel-selection policies without explicit attacker modeling. Peng et al. \cite{peng2017anti} apply Q-Learning to anti-jamming, achieving significant throughput improvements, while Luo et al. \cite{luo2018anti} leverage deep RL for spectrum sharing. Rahman et al. \cite{rahman2023rlbased} propose a multi-agent game-theoretic framework for RL-based attack detection in IoT networks, demonstrating the potential of hybrid approaches. Tang et al. \cite{tang2024adaptive} show adaptive anti-jamming strategies for UAV-aided IoT communications using reinforcement learning. Li et al. \cite{li2022drl} propose a DRL-based scheme for joint frequency and power control in dynamic wireless links, showing promising anti-jamming performance under fluctuating network conditions. Hybrid approaches combining game theory and RL have emerged, demonstrating enhanced performance \cite{zhao2020hybrid,li2021game}.

\subsection{Comparative Analyses}
Comparisons between game-theoretic and RL-based defenses are scarce. Zhao et al. \cite{zhao2020hybrid} evaluate hybrid schemes but focus on throughput only. Comprehensive surveys \cite{survey2022jamming} highlight the need for systematic comparisons across multiple performance dimensions. Our study distinguishes itself by evaluating multiple metrics—including payoffs, detection rates, and network health—under a unified experimental setup.

%======================== 3. Methodology ========================
\section{Methodology}

\subsection{Simulation Environment}
We simulate interactions on four distinct network topologies to reflect diverse IoT deployment patterns: (i) Erd\H{o}s--R\'enyi random graphs, (ii) Star topologies, (iii) Ring networks, and (iv) Small-World graphs generated via the Watts–Strogatz model. Each topology is instantiated for three different network sizes: $n = 10$, $n = 20$, and $n = 50$ nodes. For Erd\H{o}s--R\'enyi graphs, the connectivity probability is set to $p=0.4$ to ensure moderate density.

@@

Q-values are updated using the standard temporal-difference learning rule:
\[
Q(s_t, a_t) \leftarrow Q(s_t, a_t) + \alpha \left[ r_t + \gamma \max_{a'} Q(s_{t+1}, a') - Q(s_t, a_t) \right],
\]
with learning rate $\alpha = 0.1$, discount factor $\gamma = 0.9$, and $\epsilon$-greedy exploration strategy initialized with $\epsilon_0 = 0.5$ (decaying to $\epsilon_{\min} = 0.01$ at a rate of 0.998). We build our RL framework on Python-based simulation environments aligned with recent work like Jawaddi and Ismail's cloud-edge scaling platform \cite{jawaddi2024python} and RFRL Gym by Vangaru et al. \cite{vangaru2024rfrl}, which enable modular reinforcement learning experimentation in wireless systems.

@@

All simulations were conducted on a personal laptop equipped with an AMD Ryzen 5 5600H CPU and 16 GB RAM. No parallel or GPU acceleration was used. While the defense strategies are intended for deployment in resource-constrained IoT networks, the experiments were run in a simulated environment to evaluate algorithmic performance and comparative trade-offs under controlled conditions. The feasibility of deploying lightweight Q-learning-based strategies on constrained IoT hardware has been demonstrated in prior work on wireless sensor networks \cite{savaglio2019lightweight}.

@@

\subsection{Convergence Analysis}
Figure~\ref{fig:conv_db} illustrates defender payoff trajectories over time for Q-Learning and Bayesian strategies. Q-Learning defenders show rapid early improvement due to exploratory behavior, but plateau after step 150, often converging to locally optimal policies. In contrast, Bayesian defenders exhibit a slower but more stable convergence pattern, achieving higher long-term payoff with minimal inter-trial variance. This behavior reflects the advantage of equilibrium-driven decision-making under uncertainty~\cite{zeng2023comparison}.

@@

\subsection{Boxplot Comparisons}
Figure~\ref{fig:def_box} shows the distribution of final defender payoffs at $T=300$ across defender strategies (hue indicates attacker type). Bayesian defenders exhibit tighter payoff distributions (IQR = 0.2) compared to Q-Learning (IQR = 0.35), indicating more stable and predictable behavior. Additionally, the median payoff for Bayesian strategies is higher in all attacker pairings, underscoring their robustness against adversarial variability~\cite{zeng2023comparison}.

@@

Our comparative results indicate that Bayesian game-theoretic defenses offer consistently robust performance with low variance, making them well-suited for IoT applications that require predictable security guarantees. Among all evaluated metrics, the most pronounced improvement is observed in detection rate, where Bayesian defenders outperform Q-Learning counterparts by approximately 7\%. This reliability stems from the model's use of prior beliefs and equilibrium reasoning, which ensures stability even under uncertainty~\cite{zeng2023comparison}.

However, the computational overhead of belief updates and equilibrium computation may hinder real-time deployment in larger or more dynamic networks, particularly on constrained edge devices. In contrast, Q-Learning defenders demonstrate fast initial adaptation to attacker behavior, achieving higher early-stage payoffs. Yet, their long-term performance often degrades due to convergence to local optima, as also noted in recent studies on adversarial instability in RL-based security~\cite{wang2023adversarial}.

@@

Although our simulations were conducted on a general-purpose laptop, the algorithms were intentionally designed with computational efficiency in mind. For real-world IoT deployments, further evaluation on hardware platforms such as Raspberry Pi or ESP32 would help assess energy usage, memory footprint, and real-time feasibility. This concern is particularly relevant given the adversarial challenges faced by RL agents in constrained environments~\cite{wang2023adversarial}.

@@

The findings highlight a core trade-off between stability and adaptability, suggesting that future systems may benefit from hybrid designs that combine the strengths of both approaches~\cite{khoury2020hybrid}. The modular simulator and all experimental artifacts have been made publicly available to facilitate reproducibility and further exploration.
